---
title: "GLMS: Regression for Binary and Count Data"
output: pdf_document
---

```{r setup, include=FALSE}
set.seed(04192018)
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(runjags)
library(rjags)
library(tidyverse)
```

## Dichotomous Predicted Variable
- This section focus on dichotomous predicted variables: such as whether a basketball player will get a hit or if a bird will be located in a spatial grid.
\vfill
- Traditionally, these types of methods are generally implemented with logistic regression.
\vfill
- The model can be written as:

where the logistic function is $logistic(x) = 1 / (1 + exp(-x))$.

\vfill
- *Q:* What are the three components of a GLM and what are they in this specific setting?
    1. Sampling Model:
    
\vfill

    2. Linear Combination of Predictors: 

\vfill

    3. (inverse) Link function: 
    
\vfill

- *Q*: To fit this in a Bayesian framework, we need to specify priors. What parameters require priors and what distributions would be reasonable?

\newpage

We will explore the Swiss birds dataset for this lab to construct a logistic regression model for presence of the Willow Tit. 

```{r}
swiss.birds <- read.csv('http://www.math.montana.edu/ahoegh/teaching/stat491/data/willowtit2013.csv')
kable(head(swiss.birds))
```

This dataset contains 242 sites and 6 variables:  

    - siteID, a unique identifier for the site, some were not sampled during this period 
    - elev, mean elevation of the quadrant in meters 
    - rlength, the length of the route walked by the birdwatcher, in kilometers 
    - forest, percent forest cover 
    - birds, binary variable for whether a bird is observed, 1 = yes 
    - searchDuration, time birdwatcher spent searching the site, in minutes
    
#### JAGS Code

```{r}
model.string <- 'model {
  for (i in 1:Ntotal) {
    y[i] ~ dbern(mu[i])
    mu[i] <- ilogit(beta0 + sum( beta[1:Nx] * x[i,1:Nx] ))
  }

  beta0 ~ dnorm(0,1/5^2)
  for (j in 1:Nx){
    beta[j] ~ dnorm(0, 1/5^2)
  }
}'

# Fit Model
jags.logistic <-  jags.model(textConnection(model.string), 
                  data=list(y=swiss.birds$birds, 
                            Ntotal=nrow(swiss.birds), 
                            Nx = 4, 
                            x= swiss.birds[,c('elev','rlength','forest','searchDuration')]),
                  n.chains =2, n.adapt = 5000)

update(jags.logistic, 5000)
samples <- coda.samples(jags.logistic, variable.names = c('beta','beta0'), n.iter = 20000) 

summary(samples)

```



\newpage

### Interpretation of Logistic Regression Coefficients
Recall this model can be written as: $logit(\mu) = \beta_0 + \beta_1 x_1 + \dots \beta_p x_p$.

\vfill

- When $x_i$ increases or decreases by 1 unit, then logit($\mu$) increases or decreases by $\beta_i$. 

\vfill

- The logit function for $\mu$ can be expressed as $logit(\mu) = \log(\frac{\mu}{1-\mu})$. 

\vfill

- In this setting, $\mu$ is the probability of y = 1, so $$logit(\mu) = \log(\frac{Pr[y=1]}{1-Pr[y=1]}) = \log(\frac{Pr[y=1]}{Pr[y=0]})$$ 

\vfill

- This ratio: $\frac{Pr[y=1]}{Pr[y=0]}$ is known as the odds, so $logit(\mu)$ is the log-odds of an outcome in favor of 1 rather than 0.

\vfill

- Suppose the logistic regression for the Swiss birds had the following coefficients: $\beta_0 = -7$, $\beta_{elev} = .002$, and $\beta_{forest} = .06$.
    \vfill
    - Compute the probability for observing a bird in a quadrant with: elevation = 1500 meters and forest cover of 60 %. $1 / (1 + exp(- (-7 + .002 * 1500 + .06 * 60)))$ = `r 1 / (1 + exp(- (-7 + .002 * 1500 + .06 * 60)))` 
    
    \vfill
    - How do we interpret the meaning for the coefficient $\beta_{forest}$? Each unit increases the log-odds by .06. So for instance, if the forest cover was 70% rather than the 60% above the log-odds of observing a bird would increase by 0.6. 
    - The log-odds are different than a probability as in this case (conditional on the elevation) the probability increases to:
    $1 / (1 + exp(- (-6 + .002 * 1500 + .06 * 70)))$ = `r 1 / (1 + exp(- (-6 + .002 * 1500 + .06 * 70)))` 

\vfill

- Note that an unit increase in the log-odds does not have a unit increase in the probability.

\newpage
    
## Count Predicted Variable
- Now reconsider the willow tit dataset and consider modeling not just the presence / absence of birds, but directly modeling the number of birds observed in each spatial region.

```{r, fig.align='center', fig.width=4, height=3, echo = F}
birds <- read.csv('http://math.montana.edu/ahoegh/teaching/stat491/data/willowtit2013_count.csv')
head(birds)
ggplot(aes(bird.count), data=birds) + geom_bar() + ylab('Frequency') + xlab('Number of Birds Observed') + 
  ggtitle('Bird Counts of Willow Tit')
```
\vfill

1. Sampling Model: In general the Poisson model will be used as a sampling model for count data:
    -$y|\mu \sim$

\vfill

2. Linear Combination of Predictors: The same principles apply here as the other GLM settings.

\vfill

3. Link Function: 

\vfill
\vfill
